# Reproducible Research: Peer Assessment 1

Libraries
```{r, echo = TRUE}
library(knitr)
library(xtable)
library(plyr)
library(ggplot2)
library(scales)

options(scipen=999)

```

Set global settings for this assignment
```{r global settings, echo= T}
opts_chunk$set(echo= T)
Sys.setlocale("LC_TIME", "English")
```


## Loading and preprocessing the data  

Unzip the file
```{r unzip file}
if(!file.exists('Datasets'))
    dir.create('Datasets')
unzip('./activity.zip',exdir = 'Datasets')
```

Our file looks like this  
```{r check data, results='asis'}
file_path <- 'Datasets/activity.csv'
first_lines <- as.data.frame(readLines(file_path,n=10))
names(first_lines) <- 'file'
xt <- xtable(first_lines)
print(xt,type = 'html')
```
As expected, being a .csv file, the file is  
- character separated, *comma*  
- as a header  

Looking at the file and given:  
- **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
- **date**: The date on which the measurement was taken in YYYY-MM-DD format  
- **interval**: Identifier for the 5-minute interval in which measurement was taken  

Reading the data without the feature type pre-definition, we obtain the following :  
```{r read data, results='asis'}
dataset <- read.csv(file_path, header = T, sep = ',')
ds_classes <- sapply(dataset,class)
tmp <- data.frame(names(dataset),ds_classes)
names(tmp) <- c('Features','Type')
print(xtable(tmp),include.rownames=F , type = 'html')
```
Thus the *date* variable type has to be changed to a *Date* object type
```{r adjust feature type}
dataset$date <- as.POSIXct(dataset$date, format = '%Y-%m-%d')
```


## What is mean total number of steps taken per day?

```{r, warning=FALSE}
ds <- ddply(dataset, .(date), transform, totalstepsday = sum(steps, na.rm=T))

# using transform so I will use complete.cases() function
ds <- ds[complete.cases(ds),]
bin_width <- round(range(ds$totalstepsday) / 30)[2]
ggplot(ds) + 
    geom_histogram(aes(x = totalstepsday),colour = "darkgreen", fill = "white", binwidth = bin_width)  +
    xlab(label = 'Steps') +
    ylab(label = 'Frequency') +
    scale_x_continuous(breaks = pretty_breaks(n=10)) +
    scale_y_continuous(breaks = pretty_breaks(n=10))

mean_total_steps_per_day <- mean(ds$totalstepsday, na.rm = T)
median_total_steps_per_day <- median(ds$totalstepsday, na.rm = T)
```

The **mean** of the total number of steps taken per day is `r round(mean_total_steps_per_day,digits=2)` and the median is `r round(median_total_steps_per_day,digits=2)`.  

## What is the average daily activity pattern?
```{r}
ds_steps_per_interval <- ddply(dataset, .(interval), summarize, steps = mean(steps, na.rm=T))
ggplot(data =  ds_steps_per_interval) + 
    geom_line(aes(x=interval, y = steps )) +
    ylab('Avg Number of Speps')

int_maximum_number_steps <- ds_steps_per_interval[ ds_steps_per_interval$steps == max(ds_steps_per_interval$steps), 'interval']
```

The 5-minute interval that, on average across all the days in the dataset, contains the **maximum number of steps** is `r int_maximum_number_steps`.  

## Imputing missing values

```{r}
number_missing_rows <- sum(is.na(dataset$steps))
``` 

1. The total number of missing values in the dataset is `r number_missing_rows`.  
2. The strategy chosen to fill the missing values was the mean for the respective 5 minute interval.

```{r}
dataset_no_na <- dataset
for (intv in ds_steps_per_interval$interval)
    dataset_no_na[is.na(dataset_no_na$steps) & dataset_no_na$interval == intv, 'steps'] <- ds_steps_per_interval[ds_steps_per_interval$interval == intv,'steps']


```

3. The new dataset is named *dataset_no_na*


```{r}
ds_no_na <- ddply(dataset_no_na, .(date), transform, totalstepsday = sum(steps, na.rm=T))
ggplot(ds_no_na) + 
    geom_histogram(aes(x = totalstepsday),colour = "darkgreen", fill = "white", binwidth = bin_width)  +
    xlab(label = 'Steps') +
    ylab(label = 'Frequency') +
    scale_x_continuous(breaks = pretty_breaks(n=10)) +
    scale_y_continuous(breaks = pretty_breaks(n=10))

mean_total_steps_per_day_no_na <- mean(ds_no_na$totalstepsday, na.rm = T)
median_total_steps_per_day_no_na <- median(ds_no_na$totalstepsday, na.rm = T)

```` 
4. The **mean** of the total number of steps taken per day is `r round(mean_total_steps_per_day_no_na,digits=2)` and the **median** is `r round(median_total_steps_per_day_no_na,digits=2)`.  
In my case, the added values smoothed the distribution since I've used the average for each respective 5-minute interval. Thus, the mean and the median are also the same.

## Are there differences in activity patterns between weekdays and weekends?

````{r}
dataset_no_na$dayweek <- as.factor(ifelse(weekdays(dataset_no_na$date) %in% c('Sunday','Saturday'), 'weekend','weekday'))


ds_steps_per_interval_no_na <- ddply(dataset_no_na, .(interval,dayweek), summarize, steps = mean(steps, na.rm=T))
ggplot(data =  ds_steps_per_interval_no_na) + 
    geom_line(aes(x=interval, y = steps )) +
    ylab('Avg Number of Speps') +
    facet_wrap(.(dayweek), ncol = 1)

```` 

There are differences in the activity patterns between weekdays and weekends.  
On the **weekends** the user made, on average, less steps during the early hours of the day and more on the late hours in comparison with the **weekdays** average steps.
